---
title: "Final_Project"
author: "Teresa Cameron, John Light Jr, and Tyler McAfee"
date: "9/15/2019"
output: html_document
---

install.packages("readxl")
install.packages("dplyr")
install.packages("reshape2")
install.packages("ggplot2")
install.packages("maps")

```{r}
library(readxl)
states_all <- read_excel("states_all.xlsx")
```


```{r}
# Convert excel to data frame
States <- data.frame(states_all)
# Remove first column Primary_Key
States <- States[,-1]
# Filter data for years with complete information
library(dplyr)
States <- filter(States, YEAR %in% c("2003", "2005", "2007", "2009", "2011", "2013", "2015"))
# Remove territories that do not have data
States <- States[-359:-409,]
# Remove District of Columbia
States <- States[States$STATE != "DISTRICT_OF_COLUMBIA",]
# Fix column names
NamesOfColumns <- c("State_Name", "Year", "Total_Enrollment", "Total_Revenue", "Federal_Revenue", "State_Revenue", "Local_Revenue",
                    "Total_Expenditure", "Instruction_Expenditure", "Support_Service_Expenditure", "Other_Expenditure",
                    "Capital_Outlay_Expenditure", "PK_Enrollment", "KG_Enrollment", "4th_Grade_Enrollment", "8th_Grade_Enrollment",
                    "12th_Grade_Enrollment", "1st_through_8th_Enrollment", "9th_through_12th_Enrollment", "1st_through_12th_Enrollment",
                    "Average_4th_Grade_Math_Score", "Average_8th_Grade_Math_Score", "Average_4th_Grade_Reading_Score", 
                    "Average_8th_Grade_Reading_Score")
States <- setNames(States, NamesOfColumns)
# Lowercase state names
States$State_Name <- tolower(States$State_Name)
# Remove "_" and replace with space in State Name column
States$State_Name <- gsub("_", " ", States$State_Name)
# Add total expenses per student coloumn
States$Expenditure_Per_Student <- States$Total_Expenditure / States$Total_Enrollment
# Add total local revenue per student column
States$Local_Revenue_Per_Student <- States$Local_Revenue / States$Total_Enrollment
# Add total state revenue per student column
States$State_Revenue_Per_Student <- States$State_Revenue / States$Total_Enrollment
# Add total Instruction Expenses per student column
States$Instruction_Expenditure_Per_Student <- States$Instruction_Expenditure / States$Total_Enrollment
# Add total Support Services Expenditure per student coloumn
States$Support_Service_Expenditure_Per_Student <- States$Support_Service_Expenditure / States$Total_Enrollment
# Add total Capital Expenses per student column
States$Capital_Outlay_Expenditure_Per_Student <- States$Capital_Outlay_Expenditure / States$Total_Enrollment
# Add total 4th Grade test results
States$Total_4th_Grade_Score <- States$Average_4th_Grade_Math_Score + States$Average_4th_Grade_Reading_Score
# Add total 8th Grade test results
States$Total_8th_Grade_Score <- States$Average_8th_Grade_Math_Score + States$Average_8th_Grade_Reading_Score
```
Descriptive Statistics
```{r}
summary(States)
```
Plot some descriptive statistic information

```{r}
# Create data frame of revenue source by state
revenue_distribution <- States %>%
  group_by(State_Name) %>%
  summarise(avg_federal_revenue = mean(Federal_Revenue))
revenue_distribution <- data.frame(revenue_distribution)
revenue_distribution$avg_federal_revenue <- round(revenue_distribution$avg_federal_revenue)
```  
  
```{r}
# Add State Revenue
revenue_distribution1 <- States %>%
  group_by(State_Name) %>%
  summarise(avg_state_revenue = mean(State_Revenue))
revenue_distribution1 <- data.frame(revenue_distribution1)
revenue_distribution1$avg_state_revenue <- round(revenue_distribution1$avg_state_revenue)
revenue_distribution <- inner_join(revenue_distribution, revenue_distribution1, by = "State_Name")
```    
  
```{r}
# Add Local Revenue
revenue_distribution2 <- States %>%
  group_by(State_Name) %>%
  summarise(avg_local_revenue = mean(Local_Revenue))
revenue_distribution2 <- data.frame(revenue_distribution2)
revenue_distribution2$avg_local_revenue <- round(revenue_distribution2$avg_local_revenue)
revenue_distribution <- inner_join(revenue_distribution, revenue_distribution2, by = "State_Name")
```
  
```{r}
# Add total Revenue Column
revenue_distribution$Total_Revenue <- revenue_distribution$avg_federal_revenue + revenue_distribution$avg_state_revenue +
                                      revenue_distribution$avg_local_revenue
```

```{r}
# Find Federal, State, and Local Percent Revenue
revenue_distribution$Percent_Federal <- revenue_distribution$avg_federal_revenue / revenue_distribution$Total_Revenue * 100
revenue_distribution$Percent_State <- revenue_distribution$avg_state_revenue / revenue_distribution$Total_Revenue * 100
revenue_distribution$Percent_Local <- revenue_distribution$avg_local_revenue / revenue_distribution$Total_Revenue * 100
```

```{r}
library(reshape2)
rd_long <- subset(revenue_distribution, select = c(State_Name, Percent_Federal, Percent_State, Percent_Local))
rd_long <- melt(rd_long, variable.name = "Revenue_Type",
                                  value.name = "Percent")
```

  
```{r}
library(ggplot2)
rdgraph <- ggplot() + geom_bar(aes(y=Percent, x = State_Name, fill = Revenue_Type), data = rd_long, stat="identity")
rdgraph <- rdgraph + theme(axis.text.x = element_text(angle = 90, hjust = 1)) + ggtitle("Percent Revenue Distribution by State")
rdgraph
```

```{r}
# Descriptive statistics on revenue distribution
quantile(revenue_distribution$Percent_Federal)
quantile(revenue_distribution$Percent_State)
quantile(revenue_distribution$Percent_Local)
```

  
```{r}
# Create data frame of Average Expenditure Per Student per State
 g <- States %>%
  group_by(State_Name) %>%
  summarise(avg_exp_per_student = mean(Expenditure_Per_Student))
 g <- data.frame(g)
 g$avg_exp_per_student <- round(g$avg_exp_per_student, digits = 2)
``` 
install.packages("evaluate")
install.packages("mapproj")
```{r}
# Create map of Average Expenditure Per Student
us <- map_data("state")
Exp_Per_Student_Map <- ggplot(g, aes(map_id = State_Name))
Exp_Per_Student_Map <- Exp_Per_Student_Map + geom_map(map = us,
        aes(fill=avg_exp_per_student))
Exp_Per_Student_Map <- Exp_Per_Student_Map + expand_limits(x = us$long, y = us$lat)
Exp_Per_Student_Map <- Exp_Per_Student_Map + coord_map() + ggtitle("Average Expenditure per Student by State")
Exp_Per_Student_Map
```

```{r}
# Discriptive statistics on Average Expenditure per Student
quantile(g$avg_exp_per_student)
```

```{r}
# Add Average 4th Grades Test Score by State to g
h <- States %>%
  group_by(State_Name) %>%
  summarise(avg_4th_Test_Score = round(mean(Total_4th_Grade_Score)))
 h <- data.frame(h)
g <- inner_join(g, h, by = "State_Name")
```


```{r}
# Add Average 8th Grades Test Score by State to g
i <- States %>%
  group_by(State_Name) %>%
  summarise(avg_8th_Test_Score = round(mean(Total_8th_Grade_Score)))
 i <- data.frame(i)
g <- inner_join(g, i, by = "State_Name")
```

```{r}
# Create plot showing 4th Grade test scores by state and expenditures per student by region
# Add region coloumn
g$Region <- state.region 
j <- ggplot(g, aes(x=avg_exp_per_student, y= avg_4th_Test_Score))
j <- j + geom_point(aes(color=Region), size=3)
j <- j + geom_text(aes(label=State_Name), size=3, vjust=-.75, hjust=.75)
j
```

```{r}
# Create plot showing 8th Grade test scores by state and expenditures per student by region
k <- ggplot(g, aes(x=avg_exp_per_student, y= avg_8th_Test_Score))
k <- k + geom_point(aes(color=Region), size=3)
k <- k + geom_text(aes(label=State_Name), size=3, vjust=-.75, hjust=.75)
k
```

```{r}
ggplot(g, aes(x=Region,avg_4th_Test_Score)) + geom_boxplot()
```

```{r}
ggplot(g, aes(x=Region,avg_8th_Test_Score)) + geom_boxplot()
```

```{r}
# Add Instruction Expenditure per student to g
m <- States %>%
  group_by(State_Name) %>%
  summarise(avg_Instruction_Expenditure_Per_Student = round(mean(Instruction_Expenditure_Per_Student), digits =2))
m <- data.frame(m)
g <- inner_join(g, m, by = "State_Name")
```

```{r}
# Add Support Expenditure per student to g
n <- States %>%
  group_by(State_Name) %>%
  summarise(avg_Support_Service_Expenditure_Per_Student = round(mean(Support_Service_Expenditure_Per_Student), digits =2))
n <- data.frame(n)
g <- inner_join(g, n, by = "State_Name")
```

```{r}
g <- inner_join(g, revenue_distribution, by = "State_Name")
```

```{r}
# Seperate data by Region
NERegionDF <- g %>%
  filter(Region == "Northeast")
SRegionDF <- g %>%
  filter(Region == "South")
WRegionDF <- g %>%
  filter(Region == "West")
NCRegionDF <- g %>%
  filter(Region == "North Central")
```


```{r}
# Scatterplots to see if relationship
o <- ggplot(g, aes(x=avg_exp_per_student, y=avg_4th_Test_Score)) + geom_point() + stat_smooth(method = "lm", col = "red")
o
```

```{r}
p <- ggplot(g, aes(x=avg_Instruction_Expenditure_Per_Student, y=avg_4th_Test_Score)) + geom_point() + stat_smooth(method = "lm", col = "red")
p
```

```{r}
q <- ggplot(g, aes(x=avg_Support_Service_Expenditure_Per_Student, y=avg_4th_Test_Score)) + geom_point() + stat_smooth(method = "lm", col = "red")
q
```

```{r}
r <- ggplot(g, aes(x=Percent_Local, y=avg_4th_Test_Score)) + geom_point() + stat_smooth(method = "lm", col = "red")
r
```

```{r}
s <- ggplot(g, aes(x=Percent_State, y=avg_4th_Test_Score)) + geom_point() + stat_smooth(method = "lm", col = "red")
s
```


```{r}
t <- ggplot(g, aes(x=avg_exp_per_student, y=avg_8th_Test_Score)) + geom_point() + stat_smooth(method = "lm", col = "red")
t
```

```{r}
v <- ggplot(g, aes(x=avg_Instruction_Expenditure_Per_Student, y=avg_8th_Test_Score)) + geom_point() + stat_smooth(method = "lm", col = "red")
v
```

```{r}
w <- ggplot(g, aes(x=avg_Support_Service_Expenditure_Per_Student, y=avg_8th_Test_Score)) + geom_point() + stat_smooth(method = "lm", col = "red")
w
```

```{r}
x <- ggplot(g, aes(x=Percent_Local, y=avg_8th_Test_Score)) + geom_point() + stat_smooth(method = "lm", col = "red")
x
```

```{r}
y <- ggplot(g, aes(x=Percent_State, y=avg_8th_Test_Score)) + geom_point() + stat_smooth(method = "lm", col = "red")
y
```

```{r}
# Linear models
model1 <- lm(formula = avg_4th_Test_Score ~ avg_exp_per_student + Percent_Local + Percent_State + Region, data = g)
summary(model1)
```

```{r}
model2 <- lm(formula = avg_4th_Test_Score ~ Percent_Local + Percent_State + Region, data = g)
summary(model2)
```

```{r}
model3 <- lm(formula = avg_8th_Test_Score ~ avg_exp_per_student + Percent_Local + Percent_State + Region, data = g)
summary(model3)
```

```{r}
model4 <- lm(formula = avg_8th_Test_Score ~ avg_exp_per_student + Percent_Local + Percent_State, data = NERegionDF)
summary(model4)
```

```{r}
model5 <- lm(formula = avg_4th_Test_Score ~ avg_exp_per_student + Percent_Local + Percent_State, data = NERegionDF)
summary(model5)
```

```{r}
model6 <- lm(formula = avg_8th_Test_Score ~ avg_exp_per_student + Percent_Local + Percent_State, data = NCRegionDF)
summary(model6)
```

```{r}
model7 <- lm(formula = avg_4th_Test_Score ~ avg_exp_per_student + Percent_Local + Percent_State, data = NCRegionDF)
summary(model7)
```

```{r}
model8 <- lm(formula = avg_8th_Test_Score ~ avg_exp_per_student + Percent_Local + Percent_State, data = SRegionDF)
summary(model8)
```

```{r}
model9 <- lm(formula = avg_4th_Test_Score ~ avg_exp_per_student + Percent_Local + Percent_State, data = SRegionDF)
summary(model9)
```

```{r}
model10 <- lm(formula = avg_4th_Test_Score ~ Percent_Local + Percent_State, data = SRegionDF)
summary(model10)
```

```{r}
model11 <- lm(formula = avg_8th_Test_Score ~ avg_exp_per_student + Percent_Local + Percent_State, data = WRegionDF)
summary(model11)
```

```{r}
model12 <- lm(formula = avg_4th_Test_Score ~ avg_exp_per_student + Percent_Local + Percent_State, data = WRegionDF)
summary(model12)
```

```{r}
# Create categorial variable of above average test scores for 4th grade
tm4 <- mean(g$avg_4th_Test_Score)
g$good_scores_4 <- cut(g$avg_4th_Test_Score, breaks = c(-Inf, tm4, Inf),
                                                    labels = c("0", "1"))
```

```{r}
# Create categorical variable of above average test scores for 8th grade
tm8 <- mean(g$avg_8th_Test_Score)
g$good_scores_8 <- cut(g$avg_8th_Test_Score, breaks = c(-Inf, tm8, Inf),
                                                    labels = c("0", "1"))
```

```{r}
# Create train and test data sets
set.seed(1)
nrows <- nrow(g) # number of rows in the data
sample_size <- round(0.6*nrows) # take 60%

training_index <- sample(1:nrows, size = sample_size, replace = FALSE) # random index

ts_train <- g[training_index, ] # train data set
ts_test <- g[-training_index, ] # test data set
```

install.packages("kernlab")
```{r}
# Train model with the kernlab package
set.seed(1)
library(kernlab)
model <- ksvm(good_scores_4 ~ avg_Instruction_Expenditure_Per_Student + avg_Support_Service_Expenditure_Per_Student + Percent_State + Percent_Local + Region, data = ts_train,
                    kernel = "rbfdot",
                    kpar = "automatic",
                    prob.model = TRUE)

model
```

```{r}
# Table of prediction model
pred <- predict(model, ts_test)
table(pred, ts_test$good_scores_4)
```

0 - under mean 1 - over mean
2 less than average cases predicted incorrectly
5 above average cases predicted incorrectly
10 less than average cases predicted correctly
3 above average cases predicted correctly

2 + 5 = 7 error cases
7 / 20 = 0.35 total error rate %
Accuracy rate: 65%

```{r}
# Create column in ts_test for accuracy
ts_test$Predicted <- pred
Accuracy <- c()
Accuracy <- ifelse(ts_test$good_scores_4 == ts_test$Predicted, 0, 1)
ts_test$Accuracy <- Accuracy
```

```{r}
# Create scatter plot with results of model
g1 <- ggplot(ts_test, aes(x = avg_exp_per_student, y = avg_4th_Test_Score))
g1 <- g1 + geom_point(aes(size = Accuracy, color = good_scores_4, shape = pred))
g1
```

```{r}
# Train model with the kernlab package
set.seed(2)
model1 <- ksvm(good_scores_8 ~ avg_exp_per_student + avg_Instruction_Expenditure_Per_Student + avg_Support_Service_Expenditure_Per_Student + Percent_State + Percent_Local, data = ts_train,
                    kernel = "rbfdot",
                    kpar = "automatic",
                    prob.model = TRUE)

model1
```

```{r}
set.seed(2)
pred1 <- predict(model1, ts_test)
table(pred1, ts_test$good_scores_8)
```
0 - under mean 1 - over mean
0 less than average cases predicted incorrectly
6 above average cases predicted incorrectly
12 less than average cases predicted correctly
6 above average cases predicted correctly

6 error cases
6 / 20 = 0.30 total error rate %
Accuracy rate: 70%
```{r}
# Create column in ts_test for accuracy
ts_test$Predicted <- pred
Accuracy8 <- c()
Accuracy8 <- ifelse(ts_test$good_scores_8 == ts_test$Predicted, 0, 1)
ts_test$Accuracy8 <- Accuracy8
```

```{r}
# Create scatter plot with results of model
g2 <- ggplot(ts_test, aes(x = avg_8th_Test_Score, y = avg_exp_per_student))
g2 <- g2 + geom_point(aes(size = Accuracy8, color = good_scores_8, shape = pred))
g2
```

```{r}
# Create categorial variable of above average test scores for 4th grade with South regional information
tms4 <- mean(SRegionDF$avg_4th_Test_Score)
SRegionDF$good_scores_4 <- cut(SRegionDF$avg_4th_Test_Score, breaks = c(-Inf, tms4, Inf),
                                                    labels = c("0", "1"))
```

```{r}
# Create categorical variable of above average test scores for 8th grade
tms8 <- mean(SRegionDF$avg_8th_Test_Score)
SRegionDF$good_scores_8 <- cut(SRegionDF$avg_8th_Test_Score, breaks = c(-Inf, tms8, Inf),
                                                    labels = c("0", "1"))
```

```{r}
# Create train and test data sets
set.seed(1)
nrows <- nrow(SRegionDF) # number of rows in the data
sample_size <- round(0.6*nrows) # take 60%

training_index <- sample(1:nrows, size = sample_size, replace = FALSE) # random index

ts_train <- SRegionDF[training_index, ] # train data set
ts_test <- SRegionDF[-training_index, ] # test data set
```

```{r}
# Train model with the kernlab package
library(kernlab)
model <- ksvm(good_scores_4 ~ avg_Instruction_Expenditure_Per_Student + avg_Support_Service_Expenditure_Per_Student + Percent_State + Percent_Local + Region, data = ts_train,
                    kernel = "rbfdot",
                    kpar = "automatic",
                    prob.model = TRUE)

model
```

```{r}
# Make prediction table
pred <- predict(model, ts_test)
table(pred, ts_test$good_scores_4)
```

Not enough data to make prediction on regional data.